Component response rate variation drives stability in large complex systems

### 2018-06-03 notebook/ms.Rmd
New semi-major edits to the manuscript

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA alexander.duthie@stir.ac.uk
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

<!---
Current word count: 1683
--->

@@

In 1972, May [@May1972] first demonstrated that randomly assembled systems of sufficient complexity are almost inevitably unstable given infinitesimally small perturbations. Complexity in this case is defined by the size of the system (i.e., the number of interacting components; $S$), its inter-connectivity (i.e., the probability that one component will affect another; $C$), and the variance of interaction strengths ($\sigma^{2}$)[@Allesina2012]. May's finding that the probability of local stability falls to near zero given a sufficiently high threshold of $\sigma\sqrt{SC}$ has profound consequences across multiple disciplines, raising the question of how complex systems in, e.g., ecology[@Mougi2012; @Allesina2012; @Allesina2015; @Grilli2017] or banking [@May2008; @Haldane2011; @Bardoscia2017] are predicted to persist or change.

Randomly assembled complex systems can be represented as large square matrices ($M$) with $S$ components (e.g., species[@Allesina2012] or banks[@Haldane2011]). One element of such a matrix $M_{ij}$ defines how component $j$ affects component $i$ in the system at a point of equilibrium[@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation (e.g., carrying capacity in ecological communities). Traditionally, values of off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma^{2}$; diagonal elements are set to -1[@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $M$ are negative ($\max\left(\Re(\lambda)\right) < 0$)[@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly[@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma\sqrt{SC}$[@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma^{2}$ increase.

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the grey circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[fhalf]), digits = 3)` centred at $\gamma = -1.95$ and one small circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[shalf]), digits = 3)` centred at $\gamma = -0.05$. In a, the maximum real eigenvalaue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in b $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in b but not a is stable because in b $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` was stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

@@

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from 2-32 (see Supplementary Information for different values of $\sigma$ and $C$). One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$ (note that under these conditions, $\sigma\sqrt{SC} = 1$ given $S = 25$ when $\gamma = 1$). I found that the number of stable random systems was consistently higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$, and the maximum observed $S$ for which a system was stable was $31$ given $Var(\gamma)$ versus $27$ given $\gamma = 1$ (see Supplementary Information for full results). This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

Some caution is needed when interpreting these results. First, I emphasise that $Var(\gamma)$ is not stabilising per se; that is, adding variation in $\gamma$ to a particular system $M$ does not necessarily increase the probability that the system will be stable (see Supplementary Information). Rather, systems that are observed to be stable are more likely to vary in $\gamma$, and for this $Var(\gamma)$ to be critical to their stability. This is caused by the shift in the distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes result in all $\Re(\lambda) < 0$ but might also increase $\Re(\lambda)$ values. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a simple genetic algorithm (the space of possible $\gamma$ values was too large to search exhaustively[@Hamblin2013]; see Supplementary Information). For each of 10000 random $M$, the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis of each set of $\gamma$ values was performed, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a converge criteria of minimum fitness increase between generations was satisfied. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were orders of magnitude higher than when $\gamma = 1$ (see Supporting Information). Hence, managing the response rates of system components in a targetted way can potentially facilitate the stabilisation of complex systems.

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina2012]. These include systems in which all interactions are negative (e.g., competitive networks), positive (e.g., mutualist networks), or $i$ and $j$ pairs have opposing signs (e.g., predator-prey networks). In general, competitive and mutualist networks tend to be destabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist networks does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate[@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA alexander.duthie@stir.ac.uk
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - evolution.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 1
link-citations: yes
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

<!---
Current word count: 1672
--->

@@

In 1972, May [@May1972] first demonstrated that randomly assembled systems of sufficient complexity are almost inevitably unstable given infinitesimally small perturbations. Complexity in this case is defined by the size of the system (i.e., the number of potentially interacting components; $S$), its connectance (i.e., the probability that one component will interact with another; $C$), and the variance of interaction strengths ($\sigma^{2}$)[@Allesina2012]. May's finding that the probability of local stability falls to near zero given a sufficiently high threshold of $\sigma\sqrt{SC}$ has profound consequences across multiple disciplines, raising the question of how complex systems in, e.g., ecology[@Mougi2012; @Allesina2012; @Allesina2015; @Grilli2017] or banking [@May2008; @Haldane2011; @Bardoscia2017] are predicted to persist or change.

Randomly assembled complex systems can be represented as large square matrices ($M$) with $S$ components (e.g., species[@Allesina2012] or banks[@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium[@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation (e.g., carrying capacity in ecological communities). Traditionally, values of off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma^{2}$; diagonal elements are set to -1[@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $M$ are negative ($\max\left(\Re(\lambda)\right) < 0$)[@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly[@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma\sqrt{SC}$[@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma^{2}$ increase.

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the grey circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[fhalf]), digits = 3)` centred at $\gamma = -1.95$ and one small circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[shalf]), digits = 3)` centred at $\gamma = -0.05$. In a, the maximum real eigenvalue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in b $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in b but not a is stable because in b $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` was stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

@@

To investigate the effect of $Var(\gamma)$ on system stability, I simulated random $M$ matrices at $\sigma = 0.4$ and $C = 1$ across $S$ ranging from 2-32. One million $M$ were simulated for each $S$, and the stability of $M$ was assessed given $\gamma = 1$ versus $\gamma \sim \mathcal{U}(0, 2)$. I found that the number of stable random systems was consistently higher given $Var(\gamma)$ than when $\gamma = 1$ (Fig. 3), and that the difference between the probabilities of observing a stable system increased with an increase in $S$; i.e., the potential for $Var(\gamma)$ to drive stability increased with system complexity. For the highest values of $S$, nearly all systems that were stable given $Var(\gamma)$ would not have been stable given $\gamma = 1$ (see Supplementary Information for full results). This suggests that the stability of large systems might be dependent upon variation in the response rate of their individual components, meaning that factors such as generation time (in ecological networks), transaction speed (in economic networks), or communication speed (in social networks) needs to be considered when investigating the stability of complex systems.

It is important to point out that $Var(\gamma)$ is not stabilising per se; that is, adding variation in $\gamma$ to a particular system $M$ does not necessarily increase the probability that the system will be stable (see Supplementary Information). Rather, systems that are observed to be stable are more likely to vary in $\gamma$, and for this $Var(\gamma)$ to be critical to their stability. This is caused by the shift in the distribution of eigenvalues that occurs by introducing $Var(\gamma)$ (Fig. 1b, 2b), which can sometimes result in all $\Re(\lambda) < 0$ but might also increase $\Re(\lambda)$ values. To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm (the space of possible $\gamma$ values was too large to search exhaustively[@Hamblin2013]; see Supplementary Information). For each of 10000 random $M$, the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis was performed on $M$ using each set of $\gamma$ values, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a convergence criteria of minimum fitness increase between generations was satisfied. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were orders of magnitude higher than when $\gamma = 1$ (see Supporting Information), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $M$ elements. Hence, managing the response rates of system components in a targetted way can potentially facilitate the stabilisation of complex systems through a reduction in dimensionality.

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina2012]. These include systems in which all interactions are negative (competitive networks), positive (mutualist networks), or $i$ and $j$ pairs have opposing signs (predator-prey networks). In general, competitive and mutualist networks tend to be destabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist networks does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate[@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

### 2018-06-03 notebook/ms.tex
Add more notebook files; final draft edits
New semi-major edits to the manuscript

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.549 and 0.557 for a and b, respectively.

@@

To investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from 2-32 (see Supplementary Information for different
values of \(\sigma\) and \(C\)). One million \(M\) were simulated for
each \(S\), and the stability of \(M\) was assessed given \(\gamma = 1\)
versus \(\gamma \sim \mathcal{U}(0, 2)\) (note that under these
conditions, \(\sigma\sqrt{SC} = 1\) given \(S = 25\) when
\(\gamma = 1\)). I found that the number of stable random systems was
consistently higher given \(Var(\gamma)\) than when \(\gamma = 1\) (Fig.
3), and that the difference between the probabilities of observing a
stable system increased with an increase in \(S\); i.e., the potential
for \(Var(\gamma)\) to drive stability increased with system complexity.
For the highest values of \(S\), nearly all systems that were stable
given \(Var(\gamma)\) would not have been stable given \(\gamma = 1\),
and the maximum observed \(S\) for which a system was stable was \(31\)
given \(Var(\gamma)\) versus \(27\) given \(\gamma = 1\) (see
Supplementary Information for full results). This suggests that the
stability of large systems might be dependent upon variation in the
response rate of their individual components, meaning that factors such
as generation time (in ecological networks), transaction speed (in
economic networks), or communication speed (in social networks) needs to
be considered when investigating the stability of complex systems.

Some caution is needed when interpreting these results. First, I
emphasise that \(Var(\gamma)\) is not stabilising per se; that is,
adding variation in \(\gamma\) to a particular system \(M\) does not
necessarily increase the probability that the system will be stable (see
Supplementary Information). Rather, systems that are observed to be
stable are more likely to vary in \(\gamma\), and for this
\(Var(\gamma)\) to be critical to their stability. This is caused by the
shift in the distribution of eigenvalues that occurs by introducing
\(Var(\gamma)\) (Fig. 1b, 2b), which can sometimes result in all
\(\Re(\lambda) < 0\) but might also increase \(\Re(\lambda)\) values. To
further investigate the potential of \(Var(\gamma)\) to be stabilising,
I used a simple genetic algorithm (the space of possible \(\gamma\)
values was too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{18}};
see Supplementary Information). For each of 10000 random \(M\), the
genetic algorithm initialised 1000 different sets of
\(\gamma \sim \mathcal{U}(0, 2)\) values of size \(S\). Eigenanalysis of
each set of \(\gamma\) values was performed, and the 20 sets with the
lowest \(\max\left(\Re(\lambda)\right)\) each produced 50 offspring with
subsequent mutation and crossover between the resulting new population
of 1000 \(\gamma\) sets. The genetic algorithm terminated if a stable
\(M\) was found, 20 generations occurred, or a converge criteria of
minimum fitness increase between generations was satisfied. Across
\(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were found that
resulted in stable systems with probabilities that were orders of
magnitude higher than when \(\gamma = 1\) (see Supporting Information).
Hence, managing the response rates of system components in a targetted
way can potentially facilitate the stabilisation of complex systems.

I have focused broadly on random complex systems, but it is also
worthwhile to consider more restricted interactions such as those of
specific ecological
networks\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. These
include systems in which all interactions are negative (e.g.,
competitive networks), positive (e.g., mutualist networks), or \(i\) and
\(j\) pairs have opposing signs (e.g., predator-prey networks). In
general, competitive and mutualist networks tend to be destabilising,
and predator-prey network tend to be
stabilising\textsuperscript{\protect\hyperlink{ref-Allesina2011}{19}}.
When \(Var(\gamma)\) is applied to each, the proportion of stable
competitive and predator-prey networks increases, but the proportion of
stable mutualist networks does not (see Supplementary Information).
Additionally, when each component of \(M\) is interpreted as a unique
species and given a random intrinsic growth
rate\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}},
feasibility is not increased by \(Var(\gamma)\), suggesting that
variation in species generation time might be unlikely to drive
stability in purely multi-species networks (see Supplementary
Information).

>>>>

\textbf{Figure 2: Distributions of eigenvalues before (a) and after (b)
introducing variation in component response rate
(\(\boldsymbol{\gamma}\)) in complex systems.} Each panel show the same
system where \(S = 1000\), \(C = 1\), and \(\sigma = 0.4\). \textbf{a.}
Eigenvalues plotted in the absence of \(Var(\gamma)\) where
\(E[\gamma] = 1\), versus \textbf{b.} eigenvalues plotted given
\(\gamma \sim \mathcal{U}(0, 2)\), which increases the variance of
interaction strengths (\(\sigma^{2}\)) but clusters eigenvalues toward
the distribution's centre (-1, 0). Black and red elipses in both panels
show the circle centred on the distribution in panels a and b,
respectively, which have a radius of \(\sigma \sqrt{SC}\). Proportions
of \(\Re(\lambda) < 0\) are 0.549 and 0.547 for a and b, respectively.

@@

To investigate the effect of \(Var(\gamma)\) on system stability, I
simulated random \(M\) matrices at \(\sigma = 0.4\) and \(C = 1\) across
\(S\) ranging from 2-32. One million \(M\) were simulated for each
\(S\), and the stability of \(M\) was assessed given \(\gamma = 1\)
versus \(\gamma \sim \mathcal{U}(0, 2)\). I found that the number of
stable random systems was consistently higher given \(Var(\gamma)\) than
when \(\gamma = 1\) (Fig. 3), and that the difference between the
probabilities of observing a stable system increased with an increase in
\(S\); i.e., the potential for \(Var(\gamma)\) to drive stability
increased with system complexity. For the highest values of \(S\),
nearly all systems that were stable given \(Var(\gamma)\) would not have
been stable given \(\gamma = 1\) (see Supplementary Information for full
results). This suggests that the stability of large systems might be
dependent upon variation in the response rate of their individual
components, meaning that factors such as generation time (in ecological
networks), transaction speed (in economic networks), or communication
speed (in social networks) needs to be considered when investigating the
stability of complex systems.

It is important to point out that \(Var(\gamma)\) is not stabilising per
se; that is, adding variation in \(\gamma\) to a particular system \(M\)
does not necessarily increase the probability that the system will be
stable (see Supplementary Information). Rather, systems that are
observed to be stable are more likely to vary in \(\gamma\), and for
this \(Var(\gamma)\) to be critical to their stability. This is caused
by the shift in the distribution of eigenvalues that occurs by
introducing \(Var(\gamma)\) (Fig. 1b, 2b), which can sometimes result in
all \(\Re(\lambda) < 0\) but might also increase \(\Re(\lambda)\)
values. To further investigate the potential of \(Var(\gamma)\) to be
stabilising, I used a genetic algorithm (the space of possible
\(\gamma\) values was too large to search
exhaustively\textsuperscript{\protect\hyperlink{ref-Hamblin2013}{18}};
see Supplementary Information). For each of 10000 random \(M\), the
genetic algorithm initialised 1000 different sets of
\(\gamma \sim \mathcal{U}(0, 2)\) values of size \(S\). Eigenanalysis
was performed on \(M\) using each set of \(\gamma\) values, and the 20
sets with the lowest \(\max\left(\Re(\lambda)\right)\) each produced 50
offspring with subsequent mutation and crossover between the resulting
new population of 1000 \(\gamma\) sets. The genetic algorithm terminated
if a stable \(M\) was found, 20 generations occurred, or a convergence
criteria of minimum fitness increase between generations was satisfied.
Across \(S = \{2, 3, ..., 39, 40\}\), sets of \(\gamma\) values were
found that resulted in stable systems with probabilities that were
orders of magnitude higher than when \(\gamma = 1\) (see Supporting
Information), meaning that stability could often be achieved by
manipulating \(S\) \(\gamma\) values rather than \(S \times S\) \(M\)
elements. Hence, managing the response rates of system components in a
targetted way can potentially facilitate the stabilisation of complex
systems through a reduction in dimensionality.

I have focused broadly on random complex systems, but it is also
worthwhile to consider more restricted interactions such as those of
specific ecological
networks\textsuperscript{\protect\hyperlink{ref-Allesina2012}{2}}. These
include systems in which all interactions are negative (competitive
networks), positive (mutualist networks), or \(i\) and \(j\) pairs have
opposing signs (predator-prey networks). In general, competitive and
mutualist networks tend to be destabilising, and predator-prey network
tend to be
stabilising\textsuperscript{\protect\hyperlink{ref-Allesina2011}{19}}.
When \(Var(\gamma)\) is applied to each, the proportion of stable
competitive and predator-prey networks increases, but the proportion of
stable mutualist networks does not (see Supplementary Information).
Additionally, when each component of \(M\) is interpreted as a unique
species and given a random intrinsic growth
rate\textsuperscript{\protect\hyperlink{ref-Dougoud2018}{20}},
feasibility is not increased by \(Var(\gamma)\), suggesting that
variation in species generation time might be unlikely to drive
stability in purely multi-species networks (see Supplementary
Information).
